# Exploratory Data Analysis {#eda}

## Introduction

Exploratory Data Analysis (EDA) is a crucial first step in analyzing data. It provides a summary of the main characteristics of datasets and helps identify patterns, outliers, and relationships.


## Summary Statistics in R

One of the simplest ways to start analyzing data is to generate summary statistics.

```{r}
# Simple summary of built-in dataset
summary(mtcars)
```



## Data Handling

### Loading and Inspecting Data  

EDA often begins by loading the dataset and inspecting its structure.

```{r}
# Load CSV data
url <- "https://people.sc.fsu.edu/~jburkardt/data/csv/hw_200.csv"
data <- read.csv(url)

# View first rows of the data
head(data)
```



## Data Manipulation  

### Subsetting Data  

You can select specific columns or rows based on logical operations.

```{r}
# Select rows where mpg is greater than 20
subset_mpg <- mtcars[mtcars$mpg > 20, ]
head(subset_mpg)
```



### Using `dplyr` for Data Manipulation  

The `dplyr` package makes data manipulation efficient and readable.

```{r}
# Load the dplyr package
library(dplyr)

# Filter and summarize data
mtcars %>%
  filter(mpg > 20) %>%
  summarise(avg_hp = mean(hp), count = n())
```



## Visualization with Base R  

Visualizations are essential for gaining insights quickly.

```{r}
# Simple scatter plot
carsplot <- ggplot()+
  geom_point(data=mtcars, aes(mpg,hp))+
  theme_bw()
carsplot
```



## Final Thoughts  

- EDA is critical for understanding your data.  
- Data handling ensures clean datasets.  
- Data manipulation prepares data for in-depth analysis.  
- Visualization helps reveal insights quickly.



## References  

[Basic R](https://psyteachr.github.io/introdataviz/introduction.html)

[Data Explorer](https://cran.r-project.org/web/packages/DataExplorer/vignettes/dataexplorer-intro.html)

[EMSE 4572](https://eda.seas.gwu.edu/2022-Fall/class/1-getting-started.html)
[CleanData](https://www.jumpingrivers.com/blog/best-practices-data-cleaning-r/)