# Exploratory Data Analysis 

Exploratory Data Analysis (EDA) is a crucial first step in analyzing data. It provides a summary of the main characteristics of datasets and helps identify patterns, outliers, and relationships.


This section demonstrates an exploratory data analysis (EDA) of the **Palmer Penguins Dataset** using both base R and Tidyverse. It includes cleaning, descriptive statistics, and relationships between variables.

```{r echo =FALSE}
knitr::opts_chunk$set(message = FALSE,
                      eval = TRUE,
                      warning = FALSE,
                      fig.align = 'center',
                      dev = 'jpeg',
                      dpi = 300, 
                      fig.align='center')
#XQuartz is a mess, put this in your onload to default to cairo instead
options(bitmapType = "cairo") 
# (https://github.com/tidyverse/ggplot2/issues/2655)
# Lo mapas se hacen mas rapido
```

## Load Necessary Packages

```{r}
if (!require("palmerpenguins")) install.packages("palmerpenguins")
if (!require("tidyverse")) install.packages("tidyverse")

library(palmerpenguins)
library(tidyverse)
```

## Load and Clean Data

```{r}
penguins <- palmerpenguins::penguins
penguins_clean <- drop_na(penguins)

```

## Basic Data Overview

```{r}
print(head(penguins_clean))
```


```{r}
str(penguins_clean)
```


```{r}
summary(select(penguins_clean, where(is.numeric)))
```

## Measures of Central Tendency

```{r}
mean_flipper <- mean(penguins_clean$flipper_length_mm)
median_flipper <- median(penguins_clean$flipper_length_mm)
mean_flipper
median_flipper
```

## Measures of Spread

```{r}
range_body_mass <- range(penguins_clean$body_mass_g)
var_body_mass <- var(penguins_clean$body_mass_g)
sd_body_mass <- sd(penguins_clean$body_mass_g)

cat("\nRange of body mass:", range_body_mass)
cat("\nVariance of body mass:", var_body_mass)
cat("\nStandard deviation of body mass:", sd_body_mass)
```

## Frequency Tables for Categorical Data

```{r}
cat("\nFrequency of species:\n")
print(table(penguins_clean$species))
```

## Relationships Between Variables

```{r}
cov_bill <- cov(penguins_clean$bill_length_mm, penguins_clean$bill_depth_mm)
cor_bill <- cor(penguins_clean$bill_length_mm, penguins_clean$bill_depth_mm)
```

## Outlier Detection

```{r}
q1 <- quantile(penguins_clean$body_mass_g, 0.25)
q3 <- quantile(penguins_clean$body_mass_g, 0.75)
iqr_value <- q3 - q1
lower_bound <- q1 - 1.5 * iqr_value
upper_bound <- q3 + 1.5 * iqr_value

outliers <- penguins_clean %>% 
  filter(body_mass_g < lower_bound | body_mass_g > upper_bound) %>% 
  select(species, body_mass_g)

cat("\nOutliers in body mass:\n")
print(outliers)
```

## Comparison Between Groups

```{r}
penguins_clean %>% 
  filter(species %in% c("Adelie", "Gentoo")) %>%
  group_by(species) %>%
  summarise(
    mean_body_mass = mean(body_mass_g),
    sd_body_mass = sd(body_mass_g)
  )
```

## Statistical Test: t-Test

```{r}
penguins_subset <- penguins_clean %>%
  filter(species %in% c("Adelie", "Gentoo"))

t_test_result <- t.test(body_mass_g ~ species, 
                        data = penguins_subset)

cat("\nT-Test Results:\n")
print(t_test_result)
```


The null hypothesis (`H0`) assumes that the mean body mass of *Adelie* and *Gentoo* penguins is the same. If the p-value < 0.05, we reject `H0`, indicating a significant difference in the body mass of these species.


## One-Way ANOVA

```{r}
anova_body_mass <- aov(bill_length_mm ~ species, data = penguins_clean)
summary(anova_body_mass)
```

## Simple Linear Regression

```{r}
model <- lm(body_mass_g ~ flipper_length_mm, data = penguins_clean)
summary(model)
```

## Conclusion

This script demonstrates exploratory data analysis (EDA) using both Base R and Tidyverse, including cleaning, descriptive statistics, and relationships between variables. The statistical tests (t-test and ANOVA) provide insights into differences between groups, while regression analysis highlights potential predictive relationships.


## Final Thoughts  

- EDA is critical for understanding your data.  
- Data handling ensures clean datasets.  
- Data manipulation prepares data for in-depth analysis.  
- Visualization helps reveal insights quickly.



## References  

[Basic R](https://psyteachr.github.io/introdataviz/introduction.html)

[Data Explorer](https://cran.r-project.org/web/packages/DataExplorer/vignettes/dataexplorer-intro.html)

[EMSE 4572](https://eda.seas.gwu.edu/2022-Fall/class/1-getting-started.html)
[CleanData](https://www.jumpingrivers.com/blog/best-practices-data-cleaning-r/)