# Descriptive Statistics. 

Parametric and No Parametric, Linear Regression, among others

You can add parts to organize one or more book chapters together. Parts can be inserted at the top of an .Rmd file, before the first-level chapter heading in that same file. 

Add a numbered part: `# (PART) Act one {-}` (followed by `# A chapter`)

Add an unnumbered part: `# (PART\*) Act one {-}` (followed by `# A chapter`)

Add an appendix as a special kind of un-numbered part: `# (APPENDIX) Other stuff {-}` (followed by `# A chapter`). Chapters in an appendix are prepended with letters instead of numbers.




Sources in this [link](https://cienciadedatos.net/estadistica-con-r)

[here](https://www.datanalytics.com/libro_r/index.html)


[Data Types & Probability Distributions](https://fcorowe.github.io/sl/d1n1_intro.html)

LR model with [ggPredict()](https://cran.r-project.org/web/packages/ggiraphExtra/vignettes/ggPredict.html)

```{r}
data("iris")
iris
```

```{r eval=FALSE}
# What is the mean age?
mean(pirates$age)
## [1] 27

# What was the tallest pirate?
max(pirates$height)
## [1] 209

# How many pirates are there of each sex?
table(pirates$sex)
## 
## female   male  other 
##    464    490     46
```

