--- 
title: "R Basic Course"
subtitle: "This book was made with educational objetctive to improve R skill in students and professional staff"
author: "Mardones, Mauricio & Zúñiga, María José."
date: "`r Sys.Date()`"
output: pdf_document
documentclass: book
bibliography:
- book.bib
- packages.bib
url: your book url like https://mauromardones.github.io/Course_R_Statictics/
link-citations: true
linkcolor: blue
site: bookdown::bookdown_site
---


```{r setup1, echo=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

# About Course_R {-}

Description and educational material for introductory course on R, descriptive statistics, and visualizations with applications in fisheries.

This course has these sections;

- 1. **useR**. Intro to R and basic features. 
- 2. **Exploratory Data Analysis**. Basic concepts for data exploration. Essential statistical calculations. 
- 3. **Visualization**. Graphics and Maps using modern packages
- 4. **Efficient code**. Organization strategies.  
- 5. **Automatization and sharing** Report generation and result sharing.  
- 6. **Management of Bibliographic** sources and additional materials. 


All codes and used in this course can be founded in this repository with the structure to do this Ebook
[Repo R Course](https://github.com/MauroMardones/Course_R_Statistics)

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
