---
title: "Exploratory Analisys Data"
subtitle: "Curso R. Carrera Biología M/ Medio ambiente. Universidad de Chile."
author: "Mauricio Mardones"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    self_contained: true
    css: custom.css
---

```{r echo=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      out.width = "50%",
                      dev = 'jpeg',
                      dpi = 300,
                      tidy.opts=list(width.cutoff=40), 
                      tidy=TRUE)
```

## Exploratory Data Analysis (EDA)

### What is EDA?
- First step in analyzing data
- Summarizes main characteristics of datasets
- Helps detect patterns, outliers, and relationships

### Example: Summary Statistics in R

```{r}
# Simple summary of built-in dataset
summary(mtcars)
```

---

## Data Handling

### Loading and Inspecting Data
- Reading data from CSV files
- Checking structure and data types

```{r}
# Load CSV data
url <- "https://people.sc.fsu.edu/~jburkardt/data/csv/hw_200.csv"
data <- read.csv(url)

# View first rows of the data
head(data)
```

---

## Data Manipulation

### Subsetting Data
- Select specific columns or rows
- Use logical operations

```{r}
# Select rows where mpg is greater than 20
subset_mpg <- mtcars[mtcars$mpg > 20, ]
head(subset_mpg)
```

---

## Using `dplyr` for Data Manipulation

```{r}
# Load the dplyr package
library(dplyr)

# Filter and summarize data
mtcars %>%
  filter(mpg > 20) %>%
  summarise(avg_hp = mean(hp), count = n())
```

---

## Visualization with Base R

```{r}
# Simple scatter plot
plot(mtcars$mpg, mtcars$hp,
     xlab = "Miles per Gallon",
     ylab = "Horsepower",
     main = "MPG vs Horsepower")
```

---

## Final Thoughts
- EDA is critical for understanding your data
- Data handling ensures clean datasets
- Data manipulation prepares data for analysis
- Visualization helps reveal insights quickly
