---
title: "Class 3. Descriptive Statistics"
subtitle: "Curso R. Carrera Biología M/ Medio Ambiente. Universidad de Chile."
author: "Mauricio Mardones"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    self_contained: true
    css: custom.css
---

## Introduction  

- **Descriptive statistics** summarize and organize data.  
- They help us understand patterns and key characteristics in a dataset.  

We'll explore basic concepts using the `penguins` dataset from the `palmerpenguins` package.  

---

## The `penguins` Dataset  

```{r}  
# Load the data  
library(palmerpenguins)  
library(kableExtra)
data("penguins")  
kbl(head(penguins, n = c(4, 8)))
```  

---

## Basic Data Exploration  

```{r}  
# Check dimensions and structure  
dim(penguins)  
str(penguins)  
```  

---

## Summary Statistics  

```{r}  
# Quick statistical overview  
summary(penguins)  
```  

---

## Central Tendency  

- **Mean:** average value  
- **Median:** middle value  
- **Mode:** most frequent value  

```{r}  
mean(penguins$bill_length_mm, na.rm = TRUE)  
median(penguins$bill_length_mm, na.rm = TRUE)  
```  

---

## Variability  

- **Range:** difference between max and min  
- **Variance:** spread of the data  
- **Standard deviation:** square root of variance  

```{r}  
range(penguins$bill_length_mm, na.rm = TRUE)  
var(penguins$bill_length_mm, na.rm = TRUE)  
sd(penguins$bill_length_mm, na.rm = TRUE)  
```  

---

## Grouped Summaries  

```{r}  
# Grouped mean by species  
library(dplyr)  
penguins %>%  
  group_by(species) %>%  
  summarise(mean_bill_length = mean(bill_length_mm, na.rm = TRUE))  
```  

---

## Visualization: Histograms  

```{r}  
library(ggplot2)  
ggplot(penguins, aes(x = bill_length_mm)) +  
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +  
  labs(title = "Bill Length Distribution", x = "Bill Length (mm)")  
```  

---

## Boxplots for Group Comparison  

```{r}  
ggplot(penguins, aes(x = species, y = bill_length_mm, fill = species)) +  
  geom_boxplot() +  
  labs(title = "Bill Length by Species", y = "Bill Length (mm)")  
```  

---

## Correlation  

```{r}  
cor(penguins$bill_length_mm, penguins$bill_depth_mm, use = "complete.obs")  
```  

---

## Scatter Plot for Relationships  

```{r}  
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +  
  geom_point() +  
  labs(title = "Bill Length vs Depth", x = "Bill Length (mm)", y = "Bill Depth (mm)")  
```  

---

## Conclusion  

- Descriptive statistics are fundamental for understanding data.  
- R provides powerful tools to calculate, visualize, and interpret data patterns.  

Questions?  

```  