---
title: "Visualization techniques and map handling"
subtitle: "Curso R. Carrera Biología M/ Medio Ambiente. Universidad de Chile."
author: "Mauricio Mardones"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    self_contained: true
    css: custom.css
---


## Introduction  

- **Descriptive statistics** summarize and organize data.  
- They help us understand patterns and key characteristics in a dataset.  

We'll explore basic concepts using the `penguins` dataset from the `palmerpenguins` package.  

---

## The `penguins` Dataset  

```{r}  
# Load the data  
library(palmerpenguins)  
data("penguins")  
head(penguins)  
```  

---

## Basic Data Exploration  

```{r}  
# Check dimensions and structure  
dim(penguins)  
str(penguins)  
```  

---

## Summary Statistics  

```{r}  
# Quick statistical overview  
summary(penguins)  
```  

---

## Central Tendency  

- **Mean:** average value  
- **Median:** middle value  
- **Mode:** most frequent value  

```{r}  
mean(penguins$bill_length_mm, na.rm = TRUE)  
median(penguins$bill_length_mm, na.rm = TRUE)  
```  

---

## Variability  

- **Range:** difference between max and min  
- **Variance:** spread of the data  
- **Standard deviation:** square root of variance  

```{r}  
range(penguins$bill_length_mm, na.rm = TRUE)  
var(penguins$bill_length_mm, na.rm = TRUE)  
sd(penguins$bill_length_mm, na.rm = TRUE)  
```  

---

## Grouped Summaries  

```{r}  
# Grouped mean by species  
library(dplyr)  
penguins %>%  
  group_by(species) %>%  
  summarise(mean_bill_length = mean(bill_length_mm, na.rm = TRUE))  
```  

---

## Visualization: Histograms  

```{r}  
library(ggplot2)  
ggplot(penguins, aes(x = bill_length_mm)) +  
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +  
  labs(title = "Bill Length Distribution", x = "Bill Length (mm)")  
```  

---

## Boxplots for Group Comparison  

```{r}  
ggplot(penguins, aes(x = species, y = bill_length_mm, fill = species)) +  
  geom_boxplot() +  
  labs(title = "Bill Length by Species", y = "Bill Length (mm)")  
```  

---

## Correlation  

```{r}  
cor(penguins$bill_length_mm, penguins$bill_depth_mm, use = "complete.obs")  
```  

---

## Scatter Plot for Relationships  

```{r}  
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +  
  geom_point() +  
  labs(title = "Bill Length vs Depth", x = "Bill Length (mm)", y = "Bill Depth (mm)")  
```  

---

## Conclusion  

- Descriptive statistics are fundamental for understanding data.  
- R provides powerful tools to calculate, visualize, and interpret data patterns.  

Questions?  

```  
## Introduction  

- Visualizing spatial data helps communicate geographic patterns.  
- We will explore key visualization techniques using the `sf` package.  
- Example: Maps focused on Chilean administrative divisions.  

---

## Required Libraries  

```{r}  
library(sf)  
library(ggplot2)  
library(rnaturalearth)  
library(rnaturalearthdata)  
```  

---

## Loading Country Data for Chile  

```{r}  
# Download country data for Chile  
chile <- ne_states(country = "Chile", returnclass = "sf")  

# Inspect the data  
head(chile)  
```  

---

## Simple Map Plot  

```{r}  
plot(chile["name"], col = terrain.colors(nrow(chile)))  
```  

---

## Visualization with `ggplot2`  

```{r}  
ggplot(chile) +  
  geom_sf(aes(fill = name)) +  
  ggtitle("Chilean Administrative Regions") +  
  theme_minimal()  
```  

---

## Highlighting Central Regions  

```{r}  
chile_filtered <- chile[chile$name %in% c("Valparaíso", "Región Metropolitana", "O'Higgins"), ]  

ggplot(chile_filtered) +  
  geom_sf(fill = "lightblue", color = "black") +  
  ggtitle("Central Chile Map (Filtered)")  
```  

---

## Adding Cities  

```{r}  
cities <- data.frame(  
  name = c("Santiago", "Valparaíso", "Rancagua"),  
  lon = c(-70.6483, -71.6221, -70.7411),  
  lat = c(-33.4489, -33.0458, -34.1708)  
)  

cities_sf <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)  

ggplot() +  
  geom_sf(data = chile, fill = "white", color = "gray") +  
  geom_sf(data = cities_sf, color = "red", size = 3) +  
  ggtitle("Map of Major Cities in Chile")  
```  

---

## Changing Projection (CRS)  

```{r}  
chile_proj <- st_transform(chile, crs = 32719)  

ggplot(chile_proj) +  
  geom_sf(fill = "lightgreen") +  
  ggtitle("Map in UTM Zone 19S Projection")  
```  

---

## Layered Map Example  

```{r}  
ggplot() +  
  geom_sf(data = chile, fill = "lightblue") +  
  geom_sf(data = cities_sf, color = "red", size = 3) +  
  ggtitle("Map of Chile with Regions and Major Cities")  
```  

---

## Conclusion  

- `sf` and `ggplot2` are powerful tools for spatial data visualization.  
- `rnaturalearth` provides accessible boundary data for countries and regions.  
- Mapping techniques support complex geographic analyses.  

Questions?  
```  

### Notas  

- `rnaturalearth` permite obtener límites administrativos de primer nivel (regiones) para Chile.  
- El código usa el sistema de coordenadas `WGS84`, pero es posible transformar a otros sistemas (`EPSG:32719`).  
- Se pueden ajustar colores y filtros según tus preferencias.