---
title: "Visualization techniques and maps handling"
subtitle: "Curso R. Carrera Biología M/ Medio Ambiente. Universidad de Chile."
author: "Mauricio Mardones"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    css: custom.css
---



# Slide 1: Introduction

## Visualizations and Maps in R
- Importance of clear, informative visualizations
- Common mistakes to avoid
- Overview of good and bad examples

---

# Slide 2: Principles of Good Visualization
## Key Elements:
- Simplicity
- Clarity
- Relevance
- Consistency

---

# Slide 3: Good Example - Clear and Simple Plot

```{r}
library(ggplot2)
data(mpg)
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  labs(title="Engine Displacement vs. Highway Mileage")
```

- Clear labels and title
- Simple scatter plot

---

# Slide 4: Bad Example - Overloaded Plot

```{r}
ggplot(mpg, aes(x=displ, y=hwy, color=class, size=cty)) +
  geom_point() +
  labs(title="Overloaded Plot Example") +
  theme_minimal()
```
- Too much information in one plot
- Hard to interpret

---

# Slide 5: Choosing the Right Map for the Data
## Map Types:
- Choropleth maps
- Scatter plots with coordinates
- Heatmaps

---

# Slide 6: Good Example - Choropleth Map
```{r}
library(ggplot2)
library(sf)
world <- st_read(system.file("shape/nc.shp", package="sf"))
ggplot(data = world) + 
  geom_sf(aes(fill = AREA)) +
  labs(title = "Area Distribution by Region")
```
- Clear regional differences
- Proper use of color scales

---

# Slide 7: Bad Example - Poor Map Color Choices
```{r}
ggplot(data = world) + 
  geom_sf(aes(fill = AREA)) +
  scale_fill_viridis() +
  labs(title = "Bad Map Example")
```
- Color choice that doesn't differentiate well for everyone (e.g., colorblindness)

---

# Slide 8: Good Example - Using Facets in Maps
```{r}
library(dplyr)
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  facet_wrap(~class) +
  labs(title="Vehicle Class Facets")
```
- Facets make comparisons easy
- Clearly segmented data

---

# Slide 9: Bad Example - Overuse of Facets
```{r}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  facet_wrap(~class + manufacturer) +
  labs(title="Overused Facets Example")
```
- Too many facets makes it hard to extract meaning
- Can overwhelm the audience

---

# Slide 10: Mapping with Interactive Visualizations
## Using `leaflet` in R
```{r}
library(leaflet)
leaflet() %>%
  addTiles() %>%
  addMarkers(lng = -87.6298, lat = 41.8781, popup = "Chicago")
```
- Interactive maps allow user exploration
- Add custom markers and popups

---

# Slide 11: Bad Example - Static Maps for Interactive Data
```{r}
library(ggplot2)
ggplot(world, aes(fill = AREA)) + 
  geom_sf() +
  theme_void() +
  labs(title="Static Map for Interactive Data")
```
- Static map limits user interaction and exploration
- Ineffective for interactive data

---

# Slide 12: Conclusion
## Best Practices:
- Simplify your plots
- Choose the right map type
- Avoid visual clutter
- Make interactive maps when appropriate
```

This provides a presentation structure with both good and bad examples, demonstrating clear principles of effective visualization and mapping in R. You can add images or further tweak the formatting for the final output in `ioslides`.